Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-11-02 VMCAI-revised-version/final.tex
New proof only for prop~2 since the other prop use same proof

%In the case of $\pred$-basis-effective (upward) compatible WSTS, {\sc Resilience} is undecidable when $\Safe = \mathop{\downarrow} \Safe$.  This stems from the fact
%{\sc Resilience} when $\Safe= \{ q_f({\bf 0}) \}$ for some control-state $q_f$ is equivalent to
%deciding structural %termination of reset-VASS, 
%			existential halting of reset-VASS, 
%which is undecidable.

We now prove that {\sc Resilience} is undecidable for reset-VASS when 
$\Safe = \mathop{\downarrow} \Safe$.

% \mathieu{J'ai remis une vielle version que j'avais écrit avant de basculer sur le structural existential halting, je pense guarder ça en main text et détailler davantage dans l'appendix dit moi ce que tu en pense}
% In the case of $\pred$-basis-effective upward compatible WSTS, {\sc Resilience} is undecidable when $\Safe = \mathop{\downarrow} \Safe$.

@@

\begin{proof}
Let $V$ be a reset-VASS and $q_0({\bf 0})$ an initial state of $V$,
we build $V'$ from $V$ by additionnally allowing from any control-state to get back to 
$q_0({\bf 0})$
by a transition reseting every counter and
changing the control-state to $q_0$.
Then a state $q_f({\bf 0})$ is reachable from $q_0({\bf 0})$ in $N$
iff $q_f({\bf 0})$ is reachable from every element in $V'$,
which corresponds to {\sc Resilience}  for $\Safe =  \{ q_f({\bf 0}) \}$.
\end{proof}

	\iffalse
\begin{proof}
First, note that {\sc Resilience} is undecidable for a Minsky machine $M$ with at least $2$ counters when
$\Safe = \mathop{\downarrow} \Safe$,
since it is possible to have $\Safe = \{q_f(0,0)\}$. 
This stems from the fact that it is undecidable whether for a Minsky machine $M$ there exists a terminating run for every possible initial state.
\alain{et alors ? d'où sais-tu que le pb est indécidable pour safe=q,0 ? Bien sûr il l'est mais pas immédiatement puisque ce n'est pas le reachability}
Now, executions of Minsky machines can be simulated by reset-VASS~\cite{araki1976PN,dufourd1998reset} in such a way that any sequence 
% from  a state $p({\bf u})$
% to
reaching
 a state 
%
$ q({\bf 0})$ corresponds to one that reaches 
$q_f(0,0)$
 in the Minsky machine and vice versa.
 The construction rely on using pairs of counters to represent counters of the Minsky machine,
where instead of being tested for zero before a transition, one of the the counters in the pair is being reset; if the counters were zero before the transition then this was a faithful simulation, 
otherwise, there is a gap between the two counters which cannot be undone otherwise than by letting the machine backtrack.
Using the construction, which can be seen in more details in Appendix~\ref{minsk},
 we obtain that 
 {\sc Resilience} is undecidable for reset-VASS when $\Safe = \downarrow \Safe$, and hence for  $\post$-effective, $\pred$-basis-effective, 
$\omega^2$-WSTS with strong compatibility as well.\qed \end{proof}
	\fi

We will use the undecidability of the structural existential halting problem.
%
%			hence it will be also undecidable for (upward) compatible WSTS when $\Safe = \mathop{\downarrow} \Safe$.
%
%Indeed, the structural existential halting problem consists in deciding if 
%a system %terminates
%have a terminating (i.e. finite) run
% from every initial state.

% \problemx{Structural universal halting}
% {A reset-VASS $V=(Q,T)$ of dimension $d$.}
% { Do all runs for every initial state terminate ?\\}

\problemx{Structural existential halting}
%			{A reset-VASS $V=(Q,T)$ of dimension $d$.}
{A transition system $\mathscr{S}=(S,\rightarrow)$}
{ For every initial state $s_0 \in S$, does there exist a run from $s_0$ that terminates ?\\}

{\sc Structural existential halting} is undecidable for reset-VASS since termination is undecidable for deterministic counter machines. 
\alain{REFERENCE !!!!}
%			which leads to the following proposition.

% \mathieu{GOTO}

\begin{proposition}\label{indec WSTS with dcs}
{\sc Resilience} is undecidable for reset-VASS when $\Safe=\mathop{\downarrow} \Safe$, hence it is also undecidable for $\post$-effective and $\pred$-basis-effective $\omega^2$-WSTS with strong compatibility.
\end{proposition}

\begin{proof}
\nocite{araki1976PN,dufourd1998reset}
{\sc Resilience} when $\Safe= \{ q_f({\bf 0}) \}$ for some state $q_f$ is equivalent to {\sc Structural existential halting} of reset-VASS,
since a run terminates w.l.o.g. iff it reaches a terminal state of the form $q_f({\bf 0})$. \qed \end{proof}

@@

It now remains to argue why {\sc $t$-liveness} is undecidable.
Recall the
 undecidable~\cite{araki1976PN} decision problem of {\em zero-reachability} in reset-VASS% of dimension $d$%, which consists in, given a reset-VASS $V=(Q,T)$, and $p(\textbf{u}) \in Q \times \N^d$, deciding whether $\exists q \in Q ~ p(\textbf{u}) \to^* q(\textbf{0})$%
 .

@@

% 
{\sc Resilience} (and {\sc $t$-liveness}) needs a set (resp. a transition) to always remains reachable (resp. fireable), while
{\em zero-reachability} concerns itself only with the existence of at least one way to reach some state $q_f(\textbf{0})$,
which could be unreachable from other reachable states. 
%
Despite this it is possible to reduce 
the {\em zero-reachability problem} 
to
{\sc $t$-liveness}.
% The following Proposition is a variation  to reset-VASS of Theorem~5.5 in \cite{peterson1981petri} originally stated for Petri nets,
The following Proposition is a variation  to reset-VASS 
of~\cite[Theorem~5.5]{peterson1981petri} originally stated for Petri nets,
 whose proof can be seen in Appendix~\ref{appendix}%
.
\begin{proposition}\label{liveness reset}
The  zero-reachability problem can be reduced to {\sc $t$-liveness}.
\end{proposition}

% We may now prove the undecidability of the three state-resilience problems.
Since the zero-reachability problem for reset-VASS is undecidable \cite{dufourd1998reset}, the reduction implies 
{\sc $t$-liveness} is undecidable.
We deduce undecidability by reduction to {\sc State-resilience}, and straightforward reductions to the other two state resilience problems, as detailed in Appendix~\ref{appendix}.

@@

\iffalse \begin{proof}
Since the zero-reachability problem for reset-VASS is undecidable \cite{dufourd1998reset}, the previous reduction (Proposition \ref{liveness reset}) implies 
{\sc $t$-liveness} is undecidable.
Because
{\sc $t$-liveness} is undecidable in reset-VASS
and reducible to {\sc State-resilience}, we also deduce that {\sc State-resilience} is undecidable for reset-VASS, which are WSTS with strong compatibility.
Additionally, in WSTS with strong compatibility and effective $\pred$-basis, {\sc Bounded-state-resilience} is
reducible to {\sc $k$-state-resilience}:
since $\Safe=\mathop{\uparrow} \Safe$ and
$\mathscr{S}=(S,\rightarrow,\leq)$ is a WSTS with strong 
compatibility, then $\pred^{\leq n}(\Safe)= \mathop{\uparrow}~\pred^{\leq n}(\Safe)$ for all $n \in \N$,
and there exists $n_0 \in \N$ such that 
$\pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^*(\Safe) = \pred^*(\Safe)$.
We compute 
$n_0$, then iteratively check whether $k$-state-resilience 
hold for $k$ from $0$ to $n_0$.  
Furthermore, in WSTS with strong compatibility and effective $\pred$-basis,  $\Safe=\mathop{\uparrow} \Safe$, {\sc Bounded-state-resilience} is equivalent to {\sc State-resilience},
since 
$\pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^{\leq n_0}(\Safe) = {\mathop{\uparrow} \pred^*(\Safe)} = \pred^*(\Safe)$.
Hence the undecidability of {\sc Bounded-state-resilience}
and 
 {\sc $k$-state-resilience}. \qed
\end{proof} \fi

@@

We now consider the case $\Safe = \mathop{\downarrow} \Safe$.
%
Unfortunately {\sc State-resilience} is undecidable in this case. 
%\mathieu{GOTO}
This stems from undeciability of zero-reachability in reset-VASS.

@@

\begin{proof}
Let $V$ be a reset-VASS and $q_0({\bf 0})$ an initial state of $V$,
we build $V'$ from $V$ by additionnally allowing from any control-state to get back to 
$q_0({\bf 0})$
by a transition reseting every counter and
changing the control-state to $q_0$.
Then a state $q_f({\bf 0})$ is reachable from $q_0({\bf 0})$ in $N$
iff $q_f({\bf 0})$ is reachable from every element of $\post^*(q_0({\bf 0}))$ in $V'$,
which corresponds to {\sc State-resilience}  for $\Safe =  \{ q_f({\bf 0}) \}$
and $s_0 = q_0({\bf 0})$.
\end{proof}

@@

Let us recall the {\em zero-reachability} problem in (reset-)VASS.

@@

The zero-reachability problem is decidable in VASS and it is undecidable in reset-VASS \cite{araki1976PN}.
For a better representation, we will use reset Petri nets \cite{dufourd1998reset} rather than reset-VASS. 
A \emph{reset Petri net} $N = (P, T, E, R, \mu)$ consists of a finite set of {\em places} $P = \{p_1, p_2, ... p_{|P|}\}$, a finite set of {\em transitions} $T = \{t_1, t_2, ..., t_{|T|} \}$, a finite set of {\em arcs} $E \subseteq (P \times T) \cup (T \times P)$, a finite set of
{\em reset-arcs} $R \subseteq T \times P$, and an {\em initial marking} $\mu : P \to \N$.
It is well-known that Petri nets and VASS are equivalents models with regards of the decidability of reachability problems: an $d$-VASS can be encoded into a Petri net with $d+2$ places and conversely, a Petri net with $p$ places can be encoded into a $p$-VASS~\cite{DBLP:journals/siglog/Schmitz16}. The same equivalence hold between reset-VASS and reset Petri nets.

We will show that the {\em zero-reachability} problem can be reduced to {\sc $t$-liveness} in reset Petri nets; this is done by 
% "adjusting" a similar result for Petri nets (Theorem~5.5 in  \cite{peterson1981petri}).\\
"adjusting" a similar result for Petri nets~\cite[Theorem~5.5]{peterson1981petri}. \\

\noindent
\textbf{Proposition~\ref{liveness reset}.}
{\em In reset Petri nets, the zero-reachability problem can be reduced to {\sc $t$-liveness}.}\\

@@

\noindent
\textbf{Corollary}
{\em In reset-VASS, the zero-reachability problem can be reduced to {\sc $t$-liveness}.}

@@

Since the zero-reachability problem for reset-VASS is undecidable, the reduction implies {\sc $t$-liveness} is undecidable as well. This leads to the undecidability of {\sc State resilience}, {\sc Bounded-state-resilience} and
{\sc $k$-state-resilience}. 
The first follows straight from Proposition~\ref{reductions},
while the other two
follow from straightforward reductions
we nonetheless
develop in more details here.

@@

\begin{proof}
Since the zero-reachability problem for reset-VASS is undecidable \cite{dufourd1998reset}, the previous reduction (Proposition \ref{liveness reset}) implies 
{\sc $t$-liveness} is undecidable.
Because
{\sc $t$-liveness} is undecidable in reset-VASS
and reducible to {\sc State-resilience}, we also deduce that {\sc State-resilience} is undecidable for reset-VASS, which are WSTS with strong compatibility.
Additionally, in WSTS with strong compatibility and effective $\pred$-basis, {\sc Bounded-state-resilience} is
reducible to {\sc $k$-state-resilience}:
since $\Safe=\mathop{\uparrow} \Safe$ and
$\mathscr{S}=(S,\rightarrow,\leq)$ is a WSTS with strong 
compatibility, then $\pred^{\leq n}(\Safe)= \mathop{\uparrow}~\pred^{\leq n}(\Safe)$ for all $n \in \N$,
and there exists $n_0 \in \N$ such that 
$\pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^*(\Safe) = \pred^*(\Safe)$.
We compute 
$n_0$, then iteratively check whether $k$-state-resilience 
hold for $k$ from $0$ to $n_0$.  
Furthermore, in WSTS with strong compatibility and effective $\pred$-basis,  $\Safe=\mathop{\uparrow} \Safe$, {\sc Bounded-state-resilience} is equivalent to {\sc State-resilience},
since 
$\pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^{\leq n_0}(\Safe) = {\mathop{\uparrow} \pred^*(\Safe)} = \pred^*(\Safe)$.
Hence the undecidability of {\sc Bounded-state-resilience}
and 
 {\sc $k$-state-resilience}. \qed
\end{proof}

>>>>

We now prove that {\sc Resilience} is undecidable for reset-VASS when 
$\Safe = \mathop{\downarrow} \Safe$ by a reduction from the
 undecidable~\cite{araki1976PN} decision problem of {\em zero-reachability} in reset-VASS% of dimension $d$%, which consists in, given a reset-VASS $V=(Q,T)$, and $p(\textbf{u}) \in Q \times \N^d$, deciding whether $\exists q \in Q ~ p(\textbf{u}) \to^* q(\textbf{0})$%
 .

@@

\begin{proof}
Let $V$ be a reset-VASS, $q_0({\bf 0})$ an initial state of $V$,
and $q_f({\bf 0})$ a terminal state of $V$.
We build $V'$ from $V$ by additionnally allowing from any control-state to get back to 
$q_0({\bf 0})$
by a transition reseting every counter and
changing the control-state to $q_0$. Remark that the set of states 
reachable from $q_0({\bf 0})$ is the same in $V'$ and in $V$.
Then $q_f({\bf 0})$ is reachable from $q_0({\bf 0})$ in $V$
iff $q_f({\bf 0})$ is reachable from every state in $V'$,
which corresponds to {\sc Resilience}  for $\Safe =  \{ q_f({\bf 0}) \}$.
Indeed, if $q_f({\bf 0})$ is reachable from $q_0({\bf 0})$ in $V$,
then, since $q_0({\bf 0})$ is reachable from every state in $V'$,
$q_f({\bf 0})$ is reachable from every state in $V'$.
In the other direction, if $q_f({\bf 0})$ is reachable from every state in $V'$,
then in particular $q_f({\bf 0})$ is reachable from $q_0({\bf 0})$ in $V'$,
hence $q_f({\bf 0})$ is reachable from $q_0({\bf 0})$ in $V'$.
\end{proof}

@@

\noindent
It now remains to argue why {\sc $t$-liveness} is undecidable.
Recall {\sc zero-reachability} in reset-VASS
is undecidable~\cite{araki1976PN}.
\iffalse

@@

% 
\fi
{\sc Resilience} (and {\sc $t$-liveness}) needs a set (resp. a transition) to always remains reachable (resp. fireable), while
{\sc zero-reachability} concerns itself only with the existence of at least one way to reach some state $q_f(\textbf{0})$,
which could be unreachable from other reachable states. 
%
Despite this it is possible to reduce 
the {\sc zero-reachability} 
to
{\sc $t$-liveness}.
% The following Proposition is a variation  to reset-VASS of Theorem~5.5 in \cite{peterson1981petri} originally stated for Petri nets,
The following Proposition is a variation  to reset-VASS 
of~\cite[Theorem~5.5]{peterson1981petri} originally stated for Petri nets,
 whose proof can be seen in Appendix~\ref{appendix}%
.
\begin{proposition}\label{liveness reset}
{\sc zero-reachability} can be reduced to {\sc $t$-liveness}.
\end{proposition}

% We may now prove the undecidability of the three state-resilience problems.
Since {\sc zero-reachability} for reset-VASS is undecidable \cite{dufourd1998reset}, the reduction implies 
{\sc $t$-liveness} is undecidable.
We deduce undecidability by reduction to {\sc State-resilience}, and straightforward reductions to the other two state resilience problems, as detailed in Appendix~\ref{appendix}.

@@

@@

We now consider the case $\Safe = \mathop{\downarrow} \Safe$.
%
Unfortunately {\sc State-resilience} is undecidable in this case. 
%\mathieu{GOTO}
This stems from undeciability of {\sc zero-reachability} in reset-VASS,
as seen in the reduction proof of Proposition~\ref{indec WSTS with dcs}

@@

\iffalse
\begin{proof}
Let $V$ be a reset-VASS and $q_0({\bf 0})$ an initial state of $V$,
we build $V'$ from $V$ by additionnally allowing from any control-state to get back to 
$q_0({\bf 0})$
by a transition reseting every counter and
changing the control-state to $q_0$.
Then a state $q_f({\bf 0})$ is reachable from $q_0({\bf 0})$ in $N$
iff $q_f({\bf 0})$ is reachable from every element of $\post^*(q_0({\bf 0}))$ in $V'$,
which corresponds to {\sc State-resilience}  for $\Safe =  \{ q_f({\bf 0}) \}$
and $s_0 = q_0({\bf 0})$.
\end{proof}
\fi

@@

Let us recall the {\sc zero-reachability} problem in (reset-)VASS.

@@

{\sc zero-reachability} is decidable in VASS and it is undecidable in reset-VASS \cite{araki1976PN}.
For a better representation, we will use reset Petri nets \cite{dufourd1998reset} rather than reset-VASS. 
A \emph{reset Petri net} $N = (P, T, E, R, \mu)$ consists of a finite set of {\em places} $P = \{p_1, p_2, ... p_{|P|}\}$, a finite set of {\em transitions} $T = \{t_1, t_2, ..., t_{|T|} \}$, a finite set of {\em arcs} $E \subseteq (P \times T) \cup (T \times P)$, a finite set of
{\em reset-arcs} $R \subseteq T \times P$, and an {\em initial marking} $\mu : P \to \N$.
It is well-known that Petri nets and VASS are equivalents models with regards of the decidability of reachability problems: an $d$-VASS can be encoded into a Petri net with $d+2$ places and conversely, a Petri net with $p$ places can be encoded into a $p$-VASS~\cite{DBLP:journals/siglog/Schmitz16}. The same equivalence hold between reset-VASS and reset Petri nets.

We will show that  {\sc zero-reachability}  can be reduced to {\sc $t$-liveness} in reset Petri nets; this is done by 
% "adjusting" a similar result for Petri nets (Theorem~5.5 in  \cite{peterson1981petri}).\\
"adjusting" a similar result for Petri nets~\cite[Theorem~5.5]{peterson1981petri}. \\

\noindent
\textbf{Proposition~\ref{liveness reset}.}
{\em In reset Petri nets, {\sc zero-reachability} can be reduced to {\sc $t$-liveness}.}\\

@@

\noindent
\textbf{Corollary}
{\em In reset-VASS, {\sc zero-reachability} can be reduced to {\sc $t$-liveness}.}

@@

Since {\sc zero-reachability} for reset-VASS is undecidable, the reduction implies {\sc $t$-liveness} is undecidable as well. This leads to the undecidability of {\sc State resilience}, {\sc Bounded-state-resilience} and
{\sc $k$-state-resilience}. 
The first follows straight from Proposition~\ref{reductions},
while the other two
follow from straightforward reductions
we nonetheless
develop in more details here.

@@

\begin{proof}
Since {\sc zero-reachability} for reset-VASS is undecidable \cite{dufourd1998reset}, the previous reduction (Proposition \ref{liveness reset}) implies 
{\sc $t$-liveness} is undecidable.
Because
{\sc $t$-liveness} is undecidable in reset-VASS
and reducible to {\sc State-resilience}, we also deduce that {\sc State-resilience} is undecidable for reset-VASS, which are WSTS with strong compatibility.
Additionally, in WSTS with strong compatibility and effective $\pred$-basis, {\sc Bounded-state-resilience} is
reducible to {\sc $k$-state-resilience}:
since $\Safe=\mathop{\uparrow} \Safe$ and
$\mathscr{S}=(S,\rightarrow,\leq)$ is a WSTS with strong 
compatibility, then $\pred^{\leq n}(\Safe)= \mathop{\uparrow}~\pred^{\leq n}(\Safe)$ for all $n \in \N$,
and there exists $n_0 \in \N$ such that 
$\pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^*(\Safe) = \pred^*(\Safe)$.
We compute 
$n_0$, then iteratively check whether $k$-state-resilience 
hold for $k$ from $0$ to $n_0$.  
Furthermore, in WSTS with strong compatibility and effective $\pred$-basis,  $\Safe=\mathop{\uparrow} \Safe$, {\sc Bounded-state-resilience} is equivalent to {\sc State-resilience},
since 
$\pred^{\leq n_0}(\Safe) = \mathop{\uparrow} \pred^{\leq n_0}(\Safe) = {\mathop{\uparrow} \pred^*(\Safe)} = \pred^*(\Safe)$.
Hence the undecidability of {\sc Bounded-state-resilience}
and 
 {\sc $k$-state-resilience}. \qed
\end{proof}
